---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r echo = FALSE}
library(knitr)
opts_chunk$set(fig.path="figure/")
```

## Loading and preprocessing the data
```{r}
data <- read.csv("activity.csv")
data$date <- as.Date(data$date)
summary(data)
```


## What is mean total number of steps taken per day?
Total number of steps taken per day
```{r}
sum_day <- tapply(data$steps, data$date, FUN = sum, na.rm = TRUE)
head(sum_day)
```

Histogram of the total number of steps taken each day
```{r echo = T}
library(ggplot2)
ggplot()+
        geom_histogram(aes(sum_day), binwidth = 2000)+
        ggtitle("fig. 1 - Distribution of total number of steps taken each day")+
        xlab("total number of steps taken each day")
```

Mean of the total number of steps taken each day
```{r}
mean(sum_day)
```

Median of the total number of steps taken each day
```{r}
median(sum_day)
```


## What is the average daily activity pattern?

Time series of the 5-minute interval and the average number of steps taken, averaged across all days.
```{r echo = T}
mean_interval <- tapply(data$steps, data$interval, mean, na.rm = TRUE)
ggplot()+geom_line(aes(as.numeric(names(mean_interval)), mean_interval))+
        ggtitle("fig. 2 - Time series of the average number of steps taken for interval")+
        xlab("interval")+ylab("steps")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.
```{r}
names(subset(mean_interval, mean_interval == max(mean_interval)))
```

## Imputing missing values

Total number of missing values in the dataset.
```{r}
colSums(is.na(data))
```
 
Filling missing values with mean for that 5-minute interval.
```{r}
data2 <- data
for (i in 1:length(data2$steps)) {
        if (is.na(data2[i,1])) {
                data2[i,1] <- round(mean_interval[as.character(data2[i,3])])}
        }
```

Histogram of the total number of steps taken each day with filled Missing Values.
```{r echo = TRUE}
sum_day2 <- tapply(data2$steps, data2$date, FUN = sum, na.rm = TRUE)
ggplot()+
        geom_histogram(aes(sum_day2), binwidth = 2000)+
        ggtitle("fig. 3 - Distribution of total number of steps taken each day 
                with filled Missing Values")+
        xlab("total number of steps taken each day")
```

Mean of the total number of steps taken each day for with filled Missing Values.
```{r}
mean(sum_day2)
```

Median of the total number of steps taken each day for with filled Missing Values.
```{r}
median(sum_day2)
```

```{r echo = F}
options(scipen=999)
```

These values differ from the estimates done before filling the Missing Values: 

- mean of the total number of steps taken each day increased from `r round(mean(sum_day))` to `r round(mean(sum_day2))` 
- median of the total number of steps taken each day increased from `r round(median(sum_day))` to `r round(median(sum_day2))` 

Also the distribution of the total daily number of steps is different, as you can see below there are less between 0 and 2500 and more between 10.000 and 12.500.
```{r echo = TRUE}
comparison <- rbind(data.frame(sum = sum_day,missing_values = "ignored"),
                    data.frame(sum = sum_day2, missing_values = "filled"))

ggplot(comparison, aes(x=sum, fill=missing_values)) +
        geom_histogram(binwidth=2000, position="dodge")+
        ggtitle("fig. 4 - Distribution of total number of steps taken each day 
                with Ignored or Filled Missing Values")+
        xlab("total number of steps taken each day")
```

## Are there differences in activity patterns between weekdays and weekends?

Introduce a new factor variable for weekdays and weekends.
```{r echo = TRUE}
Sys.setlocale("LC_TIME", "English")
data2$weekday <- as.factor((weekdays(data2$date) == "Friday") + (weekdays(data2$date) == "Saturday"))
levels(data2$weekday) <- c("weekday", "weekend")
```

Plot the time series of the average number of steps taken for interval across all weekday days or weekend days.
```{r echo = TRUE}
ggplot(data2, aes(x = interval, y = steps)) + stat_summary(fun.y = mean, geom = "line")+ 
        facet_grid(weekday~.)+
        ggtitle("fig. 5 - Time series of the average number of steps taken for interval 
                across all weekday days or weekend days")
```